```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[**Lab #1 Markdown File**](/assignments/Lab_1.Rmd)

[**Lab Instructions**](https://dssoc.github.io/assignments/lab_instructions.html)

In this lab we will be learning the basics of using R and working with real data.

See the "Instructions" section of the [Lab Instructions](https://dssoc.github.io/assignments/lab_instructions.html) page for more information about the labs. That page also gives descriptions for the datasets we will be using.

**Required reading:**

-   [R for Data Science Chapter 4: Workflow basics](https://r4ds.had.co.nz/workflow-basics.html)
-   [R4DS Chapters 26-30: Communicating your analysis](https://r4ds.had.co.nz/communicate-intro.html) (Except Chapter 28 - we will cover visualization later.)

**Optional resources:**

-   [RDocumentation.org](%5Bdocumentation%20website%5D(https://www.rdocumentation.org/))

```{r message=FALSE, echo=FALSE}
# Load the datasets and libraries. You shouldn't need to change the URL in the `load` function.

library(tidyverse)
library(readxl)

# this will download the dataset from the course website
load(url('https://dssoc.github.io/datasets/congress.RData'))
```

<br/>

## Practice Questions

**ex1. What is the sum of the values of `some_vector`?**

```{r}
some_vector <- c(25555,342343,123123123,4234234,53243234,54324234,5421111,12312312,111231)
sum(some_vector)
```

**ex2. Compute the sum of all integers 1 through 100, inclusive.**

HINT: The solution to this problem is found in the required readings.

```{r}
sum(seq(1, 100))
```

## Questions

<br>

**1. Use the [documentation website](https://www.rdocumentation.org/) or the helper functions in R (`?function_name`) to look up the `mean` function, and describe each of its arguments and what they do in your own words. Write your answer in the code block below as a comment.**

NOTE: You can write comments in a code block by including a "\#" character at the beginning of the line.

```{r}
# There are two main implmentations of the mean function in R. The basic implmentation would look like this "mean(x,...)" where the x would be a R object which the mean will be then be taken of. This can even be called on more complex vectors but only in the more complicated implemenation where the trim=0. Now to get to the more complciation implmenation, it would look a little more like this "mean(x,trim,na.rm,...)". In this siutaiton x is still the same (an R object), but trim would be a value between 0 and 0.5 inclusive by which the value in x will be trimmed before the mean is taken. Values outside the range will be trimmed to the nearest endpoint. The na.rm checks to see if the logical operator NA should be stripped out befor ethe mean is taken and would be either TRUE or FALSE when called in the mean method to see whether the NA operators are removed or not. In both implmentations the ... would be extra arguments passed to or from other methods. 
```

<br>

**2. Extract the third element of the vector using numerical indexing.**

NOTE: the first element index starts with 1 in R.

```{r}
my_vector <- c("R","is","great")
third <- my_vector[3]
print(third)
```

<br>

**3. This question has several parts: (a) Use R code to identify the data type of `some_vector`. (b) What is the largest number in this vector? (c) How about the mean value? (d) Does this vector include the number `54324222`? Your results should be generated from the R code - please do not write the numbers into R manually.**

HINT: you'll have to research some new functions in r to do this- try Google or one of the tutorials linked on the syllabus.

HINT: check out the `%in%` operator.

```{r}
some_vector <- c(25555,342343,123123123,4234234,53243234,54324234,5421111,12312312,111231,
                     1231231,12312312,12312312,123123,898972,789872,2343,23423423,2343221,23423,
                     14444,44324222,2342341,124231111,22233345,1111233333,1231231,1231231)
# a)
print(typeof(some_vector))
# b)
print(max(some_vector))
# c)
print(mean(some_vector))
# d)
print(54324222 %in% some_vector)
```

<br>

**4. Use the `congress` dataframe for the remainder of questions in this assignment. (a) How many rows and columns does the `congress` dataframe have? (b) Use a function to show its data type. You must use R code to generate these values.**

```{r}
# a) 539 rows and 8 columns
# b)
sapply(congress, class)
```

<br>

**5. Using the browse features of R or RStudio, write the ages of Sherrod Brown (a member of congress) and Dianne Feinstein.**

```         
Your written response here.
Sherrod Brown: 72
Dianne Feinstein: 91 
```

<br>

**6. What is the average birthyear of congress members? What is the data type of the `birthyear` column?**

```{r}
# Data type: numeric
print(mean(congress$birthyear))
```

<br>

**7. What are your learning goals for this semester? Not a trick question ;)**

```         
Get a better understanding of social and cultural implications of AI/ML(and computer science as a whole). A more specific situation would be understanding how NLP can be utilized to help alleviate the increasing political polarization in the US as a whole.
```
